---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ezEDA


<!-- badges: start -->
<!-- badges: end -->

The goal of ezEDA is to enable users to create visualizations using functions based on the data analysis task rather than on plotting mechanics. It hides the details of the individual 'ggplot2' function calls and allows the user to focus on the end goal. Useful for quick preliminary explorations. ezEDA provides functions for common exploration patterns. Some of the ideas in this package are motivated by Fox (2015, ISBN:1938377052).

## Installation

You can install the released version of ezEDA from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ezEDA")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, echo = FALSE}
library(ezEDA)
## basic example code
category_tally(ggplot2::mpg, class)
```


