---
title: "ezEDA: A Task-Oriented Interface for Exploratory Data Analysis "
author: "Viswa Viswanathan"
date: "May 25, 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction
Owing to its extensive functionality and its roots in a robust grammar of graphics (Ref. Wilkerson), ggplot (ref Hadley Wickham) has become very popular among R users. Using ggplot, we develop visualizations by mapping elements of the dataset to features of the plot and build the required plot layer by layer. The ggplot package also provides the convenience function qplot that allows us to generate many common plots with a single function call.

Although exploratory analysis of data differs from dataset to dataset, we can still see some recurrent themes or patterns that apply to many situations and ezEDA aims to ease the application of these themes. With ezEDA we aim to ease the task of generating ggplot-based visualizations by allowing users to think in terms of their problem domain rather than the details of how to achieve the plot that they have in mind. This approach is particularly useful when the visualization task involves standard themes or patterns. The ezEDA package currently provides functions for twelve patterns and we aim to incorporate more in future versions.

ezEDA, allows the user to think more in terms of the problem domain and forget about the mechanics of producing the plot. The user calls a single function that represents a generic pattern and gets the visualization needed. We feel that this is especially useful when someone is actively exploring a dataset and wants to quickly generate some visualizations to get an idea about some aspects of the dataset. ezEDA does not aim to complelely hide ggplot. Instead we intend it to ease the application of a small number of common visualization patterns. By its very nature, ezEDA does not allow users to control all aspects of the plots. It uses certain defaults and allows the user to quickly generate visualizations. Once they have completed their initial explorations, it could very well be that users will then dive into ggplot and fine-tune their plots if needed.

Like ggplot and dplyr, ezEDA makes interactive use easier by allowing the use of unquoted column names.

Currently, ezEDA provides functions under five categories: 

* Tallies
* Contributions
* Relationships
* Distributions, and 
* Trends


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Basic concepts behind design of the ezEDA interface
ezEDA is partly motivated by the work of Stephen Few (ref) who described a general framework for Exploratory Data Analysis. Such a framework would be useful when we are presented with a dataset and would like to generate questions to ask of the dataset -- questions that could potentially generate interesting results or confirm what we already knew or suspected. Few's framework is driven by the observation that any dataset has columns of four types:

* categories -- categorical columns representing categories with no inherent numerical meaning
* measures -- numerical columns
* time -- columns representing time (could be date columns or just integers representing sequential time units)
* others -- columns without much significance for data exploration (like names, etc.)

Once we establish this terminology, the list below shows the patterns and the corresponding ezEDA functions:

* tallies for counts of one or two categories
    * functions: `category_tally` and `two_category_tally`
* contributions of different categories to a measure
    * functions: `category_contribution` and `two_category_contribution`
* relationships between measures
    * functions: `two_measure_relationship` and `multi_measure_relationship` 
* trends of measures
    * functions: `measure_change_over_time_wide` and `measure_change_over_time_long`
* distributions of measures
    * functions: `measure_distribution`, `measure_distribution_by_category`, `measure_distribution_by_two_categories` and `measure_distribution_over_time`

ezEDA provides functions for these tasks. The package currently has twelve functions and we hope to add more as we identify more patterns.

## Installing ezEDA
ezEDA is available under CRAN and can be installed as usual by calling the install.packages function:

```{r eval=FALSE}
install.packages("ezEDA")
```
ezEDA depends on many other packages like ggplot2, dplyr, tidyr and others and if any of those packages are not installed on your system, the above code will cause any of the missing packages to be installed as well.
## Using ezEDA
As always, it is a good idea to make the package namespace available via the library function:

```{r setup}
library(ezEDA)
```
## ezEDA functions
We discuss each of the functions below. For convenience, we have divided the functions into convenient groups.

### Tallies
The two functions in this group satisfy a common need to generate counts of categories. For example, in a dataset of students, we might want to plot the counts of students by gender; in the diamonds dataset from ggplot2, we might want to generate a plot of the number of diamonds by color, and so on.

#### category_tally
Barplot of the counts based on a category column. Often when a barplot has many bars, we run the risk of the x-axis labels running onto each other. To handle this issue, the function flips the coordinates. The first argument is the dataset to be used and the second argument is the unquoted name of the relevant category column.
```{r category_tally}
## For ggplot2::mpg, plot the tallies for the different vehicle classes
category_tally(ggplot2::mpg, class)
```

#### two_category_tally
Barplot of one category showing its conposition in terms of another category. The arguments are: the dataset to be used and the unquoted names of the two category columns.
```{r two_category_tally_1}
## For ggplot2::diamonds plot the tallies for different types of cut and clarity
two_category_tally(ggplot2::diamonds, cut, clarity)
```

Below is a variant with facets.
```{r two_category_tally_2}
## For ggplot2::diamonds, plot the tallies for different types of cut and clarity
## and show the plots for each value of the second category in a separate facet
two_category_tally(ggplot2::diamonds, cut, clarity, separate = TRUE)
```

### Contributions
The two functions in this group satisfy a common need to analyze the extent to which specific categories contribute to specific measures. For example, in a dataset of people, we might want to plot the total income by gender; in the diamonds dataset from ggplot2, we might want to generate a plot of the contribution to the price by diamonds of various kinds of cut, and so on.

#### category_contribution
Barplot of the total of a measure based on a category column. 
```{r category_contribution}
## For ggplot2::diamonds, plot the total price for each kind of cut
category_contribution(ggplot2::diamonds, cut, price)
```

#### two_category_contribution
Barplot of the total of a measure based on a category column. 
```{r two_category_contribution_1}
## For ggplot2::diamonds, plot the total price for each kind of cut while also showing 
## the contribution of each kind of claritywithin each kind of cut
two_category_contribution(ggplot2::diamonds, cut, clarity, price)
```

Below is a variant with facets.
```{r two_category_contribution_2}
## For ggplot2::diamonds, same as above, but show each color on a different facet
two_category_contribution(ggplot2::diamonds, cut, clarity, price, separate = TRUE)
```

### Relationships
The two functions in this group visualize relationships between measures.

#### two_measures_relationship
Scatterplot of the relationship between two measures with optional coloring of points by a category.
```{r two_measures_relationship_1}
## For ggplot2::mpg, plot the highway mileage against the displacement 
two_measures_relationship(ggplot2::mpg, displ, hwy)
```

Below is a variant with coloring of points by a category.
```{r two_measures_relationship_2}
## For ggplot2::mpg, plot the highway mileage against the displacement while showing 
## cars of different class through the color of the point
two_measures_relationship(ggplot2::mpg, displ, hwy, class)
```

#### multi_measures_relationship
Scatterplot matrix showing the relationships between three or more measures.  
```{r multi_measures_relationship}
## For ggplot2::mpg, plot the relationship between city mileage, highway mileage and displacement 
multi_measures_relationship(ggplot2::mpg, cty, hwy, displ)
```

### Distributions
The four functions in this group visualize distributions of measures.

#### measure_distribution
Distribution of a measure: default is histogram. 
```{r measure_distribution_1}
## For ggplot2::mpg, plot the distribution of highway mileage
measure_distribution(ggplot2::mpg, hwy)
```
Distribution of a measure: default is histogram. By default the function uses a bin width corresponding to 30 bins. Use the bwidth argument to specify the desired value for the width of each bin.
```{r measure_distribution_2}
## For ggplot2::mpg, plot the distribution of highway mileage 
measure_distribution(ggplot2::mpg, hwy, bwidth = 2)
```

Distribution of a measure. Get a boxplot instead of histogram.
```{r measure_distribution_3}
## For ggplot2::mpg, plot the distribution of highway mileage as a boxplot
measure_distribution(ggplot2::mpg, hwy, type = "box")
```
#### measure_distribution_by_category
Distribution of a measure with highlighting of different values of a single category.
```{r measure_distribution_by_category_1}
## For ggplot2::diamonds, plot the distribution of price while highlighting 
## the counts of diamonds of different cuts 
measure_distribution_by_category(ggplot2::diamonds, price, cut)
```

Distribution of a measure with highlighting of different values of a single category across facets.
```{r measure_distribution_by_category_2}
## For ggplot2::diamonds, plot the distribution of price showing the distribution
## for each kind of cut in a different facet
measure_distribution_by_category(ggplot2::diamonds, price, cut, separate = TRUE)
```
#### measure_distribution_by_two_categories
Distribution of a measure for a combination of two categories within a facet grid
```{r measure_distribution_by_two_categories}
## For ggplot2::diamonds, plot the distribution of price separately for each unique combination
## of cut and clarity
measure_distribution_by_two_categories(ggplot2::diamonds, carat, cut, clarity)
```

#### measure_distribution_over_time
STudy the change of distribution of a measure over time
```{r measure_distribution_over_time}
h1 <- round(rnorm(50, 60, 8), 0)
h2 <- round(rnorm(50, 65, 8), 0)
h3 <- round(rnorm(50, 70, 8), 0)
h <- c(h1, h2, h3)
y <- c(rep(1999, 50), rep(2000, 50), rep(2001, 50))
df <- data.frame(height = h, year = y)
measure_distribution_over_time(df, h, year)
```
### Trends
The two functions in this group simultanepusly visualize trends of up to 6 measures. Depdning on whether the data is in wide form (different measures in different columns) or long (all measure values in one column with another column to identify each measure) you can use a differemnt function.

#### measure_change_over_time_wide
Simultaneously visualize the change over time of several measures (up to 6) when the data is in wide form (see above). 
```{r measure_change_over_time_wide}
## For ggplot2::economics, plot the trend of population and number unemployed. In this dataset
## the different measures are in different columns
measure_change_over_time_wide(ggplot2::economics, date, pop, unemploy)
```
#### measure_change_over_time_long
Simultaneously visualize the change over time of several measures (up to 6) when the data is in long form (see above). 
```{r measure_change_over_time_long}
## For ggplot2::economics_long, plot the trend of population and number unemployed. In this dataset
## all measures are in the column value and the names of the measures are in the column variable
measure_change_over_time_long(ggplot2::economics_long, date, variable, value, pop, unemploy)
```
